### 2016-09-01 latex/paper.tex
Remove underlines from references
Minor edits

\section*{ABSTRACT}
The spokes of a bicycle wheel brace and stiffen its slender rim, but only as long as the pre-tension is high enough to prevent them from going slack under external loads. However, above a certain critical spoke tension, the rim will buckle into a non-planar shape. Despite its importance for wheel strength and safety, there is no formula in the literature for the maximum tension a wheel can withstand before buckling.

Here we derive such a formula. The non-dimensionalized buckling tension depends only on two parameters which are identified as the ratio of rim torsional stiffness to rim bending stiffness, and the ratio of lateral spoke bracing stiffness to rim bending stiffness. The buckled rim adopts a sinusoidal lateral profile where the number of positive peaks $n\geq 2$ is determined by these two stiffness ratios. We show that the wheel may be represented by a system of equivalent springs: two series-connected springs for the rim bending and torsion stiffness, connected in parallel with a spring for the spoke system.

@@

He did not discuss the problem of buckling under tension further, but proposed that wheel failure commonly occurs due to lateral buckling when external loads have caused spokes to lose tension. Pippard and Francis~\cite{Pippard1932d} derived an equation for lateral deflection based on an elastic foundation model, but did not discuss stability and neglected the effect of spoke tension.

The theory of bicycle wheel buckling is closely related to the theory of flexural-torsional buckling of arches. Timoshenko and Gere~\cite{Timoshenko1961a} gave a formula for the critical load for a ring with doubly-symmetric cross-section subjected to radial loads. The theory of flexural-torsional buckling of monosymmetric arches (bicycle rim cross-sections generally have only one plane of symmetry) was broadly formalized by Trahair and Papangelis~\cite{Trahair1987b} using the virtual work approach to derive the equilibrium and stability equations. Their theory has been extended to treat arches restrained by continuous~\cite{Pi2002b} or discrete~\cite{Bradford2002d} elastic supports and elastic end restraints~\cite{Guo2014b}.

@@

@@

To hold the spokes in tension, the rim must be under compression. An accurate approximation\footnote{95\% accurate for 6 spokes, $\alpha=10^{\circ}$, $\beta=10^{\circ}$, where angles $\alpha$ and $\beta$ are defined in Fig. \ref{fig:schematic}. Increasing the number of spokes only improves the estimate.} was given by Sharp~\cite{Sharp1977a} by considering a force balance on the top (or bottom) half of the rim.
	\begin{equation}\label{eq:TN}
	N_r = \frac{n_sT}{2\pi}
	\end{equation}
The spokes act like guy-wires, both supporting the rim radially and stabilizing the rim laterally. However, if the spoke tension exceeds a critical value $T_c$, the compression in the rim causes the wheel to buckle laterally, bending and twisting out of its initial plane, as shown in Fig. \ref{fig:schematic} (c).

@@

The deformed configuration of the rim is described by the displacement of the shear center $\bm{u}=u\e + v\ee + w\eee$ and the twist angle $\phi$. The in-plane curvature $\kappa_1$ remains constant during flexural-torsional buckling. The in-plane curvature, out-of-plane curvature and twist rate are
	\begin{equation}\label{eq:k1}
	\kappa_1 = 1/R
	\end{equation}
	\begin{equation}\label{eq:k2}
	\kappa_2 = u'' - \frac{\phi}{R}
	\end{equation}
	\begin{equation}\label{eq:k3}
	\kappa_3 = \phi' + \frac{1}{R} u'
	\end{equation}
The symbol $()'$ indicates a derivative with respect to arc-length $s$ along the \eee{} direction. It can easily be verified that Equations \ref{eq:k1}--\ref{eq:k3} simplify to the standard equations for straight beams in the limit $R\rightarrow \infty$. The total strain energy is decomposed into lateral bending, uniform torsion, and warping terms which are related to the curvatures.
% The curvature and twist rate produce a bending moment and twisting moment equal to
	% \begin{equation}\label{eq:M2}
	% M_2 = EI_{11} \kappa_2
	% \end{equation}
	% \begin{equation}\label{eq:M3}
	% M_3 = GJ \kappa_3 + EI_w \kappa_3''
	% \end{equation}
	\begin{equation}\label{eq:Urim}
	U_{rim} = \frac{1}{2} \int_0^{2\pi} \left( EI_{11} \kappa_2^2 + GJ \kappa_3^2 + EI_w (\kappa_3')^2 \right)\, ds
	\end{equation}
The warping energy is primarily related to the bending energy in the rim sidewalls (which act as I-beam flanges) of the rim due to twisting\footnote{See http://www.steel-insdag.org/TeachingMaterial/Chapter17.pdf for a simple treatment.}. For a single-wall rim, the energy contribution from warping can exceed the contribution from torsion.

@@

In this paper we consider bicycle wheels with slender prestressed spokes. We model a single spoke as an elastic bar pinned at each end, ensuring that the spoke force is always collinear with the spoke axis. The hub is assumed to be rigid. The linearized elongation of a spoke is given by
	\begin{equation}\label{eq:selong}
	\Delta l = \bm{u}_n\cdot \n
	\end{equation}
where $\bm{u}_n$ is the displacement vector of the spoke nipple and \n{} is a unit vector pointing along the axis of the spoke. The linearized rotation of the spoke is given by
	\begin{equation}\label{eq:srot1}
	\Omega_1 = \frac{1}{l} \bm{u}_n\cdot \nn
	\end{equation}
	\begin{equation}\label{eq:srot2}
	\Omega_2 = \frac{1}{l} \bm{u}_n\cdot \nnn
	\end{equation}
where \nn{} and \nnn{} are two orthogonal unit vectors orthogonal to \n{}, and $l$ is the spoke length. The elongation produces a net force on the rim parallel to the original spoke axis equal to
	\begin{equation}\label{eq:sF1}
	f_1 = E_sA_s\left(\frac{\Delta l}{l}\right)\n = \frac{E_sA_s}{l} \n \cdot \bm{u}_n 
	\end{equation}
The rotation of the spoke produces net forces on the rim perpendicular to the original spoke axis due to the rotation of the initial tension vector.
	\begin{equation}\label{eq:sF2}
	f_2 = T \sin{\Omega_1} \approx \frac{T}{l} \nn \cdot \bm{u}_n
	\end{equation}
	\begin{equation}\label{eq:sF3}
	f_3 = T \sin{\Omega_2} \approx \frac{T}{l} \nnn \cdot \bm{u}_n
	\end{equation}
since \n{}, \nn{}, \nnn{} are mutually orthogonal, the total force on the rim is given by $\bm{f} = \bm{k}_f \cdot \bm{u}_n$, where the spoke stiffness matrix is given by
	\begin{equation}\label{eq:kf}
	\bm{k}_f = \frac{E_sA_s}{l}\n\n + \frac{T}{l}(\nn\nn + \nnn\nnn)
	\end{equation}
The first term in Eqn.~\ref{eq:kf} is the {\bf elastic stiffness} and the second term is the {\bf tension stiffness}. The tension stiffness is the component responsible for the transverse vibrations of a guitar string, e.g. The term $\n\n$ is the dyadic product, or tensor product (also called the outer product), and is defined by the property $\n\n\cdot \bm{a} = (\n\cdot \bm{a})\n$ for any vector $\bm{a}$.

@@

\subsection{Total potential energy and buckling criterion}
The wheel becomes unstable when the increase in strain energy is exactly balanced by the virtual work for a perturbing displacement $\delta u, \delta\phi$, i.e. when the total potential energy is zero.
	\begin{equation}\label{eq:TotPot}
	\Pi = U_{rim} + U_{spokes} - V_{rim} = 0
	\end{equation}
We assume a buckled shape of the form
	\begin{equation}\label{eq:modeshape}
	\begin{split}
	u &= \delta u_n \cos{\frac{ns}{R}} \\
	\phi &= \delta\phi_n \cos{\frac{ns}{R}}
	\end{split}
	\end{equation}
\,\,\,\, where $n$ is the number of positive peaks of the lateral displacement. Substituting the buckling mode shape Eqn. \ref{eq:modeshape} into Eqns. \ref{eq:Urim}, \ref{eq:Vrim}, and \ref{eq:Us} and substituting into Eqn. \ref{eq:TotPot} yields a buckling criterion for the $n$th mode.
	\begin{multline}\label{eq:ECrit}
	\frac{\pi EI_{11}}{R^3}(n^2 \delta u_n - R\delta\phi_n)^2 + \frac{\pi n^2}{R^2}\left(GJ + \frac{EI_w}{R^2}n^2\right)(\delta u_n-R\delta\phi_n)^2 + \\
	\pi R(\bar{k}_{uu}^{el} + \bar{k}_{uu}^{el})\delta u_n^2 - N_r\frac{\pi n^2}{R}\delta u_n^2=0
	\end{multline}
The torsional stiffness and warping stiffness can be combined into an effective torsional stiffness $\widetilde{GJ} = GJ + n^2EI_w/R^2$. Equation \ref{eq:ECrit} has a quadratic form and can be written in matrix notation:
	\begin{equation}\label{eq:Qform}
	\begin{bmatrix}
	\delta u_n\\\delta\phi_n
	\end{bmatrix} \cdot
	\begin{bmatrix}
	\frac{\partial^2 \Pi}{\partial u_n^2} & \frac{\partial^2 \Pi}{\partial u_n \partial \phi_n}\\
	\frac{\partial^2 \Pi}{\partial u_n\partial\phi_n} & \frac{\partial^2 \Pi}{\partial \phi_n^2}
	\end{bmatrix} \cdot
	\begin{bmatrix}
	\delta u_n\\\delta\phi_n
	\end{bmatrix}
	=0
	\end{equation}
Non-trivial solutions of Eqn. \ref{eq:Qform} exist when the determinant of the matrix vanishes. Expanding the determinant in Eqn. \ref{eq:Qform} and substituting Eqns. \ref{eq:TN} and \ref{eq:kuu} yields a linear equation for the critical buckling tension of the $n$th mode.
	\begin{multline}\label{eq:Tc_eq}
	\left(\frac{2\pi EI_{11}}{R}+\frac{2\pi \widetilde{GJ}n^2}{R} \right) \left( \frac{2\pi \widetilde{GJ}n^2}{R^3} + \frac{2\pi EI_{11}n^4}{R^3} + 2\pi R\bar{k}_{uu}^{el} + \frac{n_sT_{cn}}{l} - \frac{n_sn^2T_{cn}}{R}\right)\\
	- \left( \frac{2\pi EI_{11}n^2}{R^2} + \frac{2\pi \widetilde{GJ}n^2}{R^2} \right)^2 = 0
	\end{multline}
Solving Eqn. \ref{eq:Tc_eq} for $T_{cn}$ yields the tension at which the $n$th mode becomes unstable. The critical buckling tension $T_c$ is the minimum $T_{cn}$ with respect to the mode number $n\geq 2$.

\section{BUCKLING DUE TO SPOKE TENSION}
The solution to Eqn. \ref{eq:Tc_eq} can be written in non-dimensional form as
	\begin{equation}\label{eq:Tc_nondim}
	\frac{T_{cn}}{T_e} = \left( \frac{\tilde{u}n^2(n^2-1)^2}{1+\tilde{\mu}n^2} + \lambda \right) \left(\frac{1}{n^2-R/l} \right)
	\end{equation}
\,\,\,\, where the ``Euler tension'' $T_e=2\pi EI_{11}/n_sR^2$ is the tension that would produce a compressive force in the rim equal to the classical buckling load for a straight beam-column of length $2\pi R$. The parameter $\tilde{\mu} = \widetilde{GJ}/EI_{11}$ is the ratio of the effective torsional stiffness to the lateral bending stiffness. The parameter $\lambda = \bar{k}_{uu}^{el}R^4/EI_{11}$ is the ratio of the spoke system stiffness $\bar{k}_{uu}^{el}R$ to the rim bending stiffness $EI_{11}/R^3$. For most wheels, $R/l\approx 1$.

@@

\subsection{Rims with low torsional stiffness}\label{sec:powerlaw_1}
Many bicycle rims are built from thin-walled open channel sections which have very low torsional stiffness $\widetilde{GJ}$ compared with their bending stiffness. Replacing the discrete mode number $n$ with a continuous variable $\bar{n}$, we adopt the following ansatz: (a) $\bar{n}^2 \gg 1$ and (b) $\tilde{\mu}\bar{n}^2 \ll 1$. Under these conditions, Eqn. \ref{eq:Tc_nondim} becomes
	\begin{equation}\label{eq:Tcn_small_mu}
	T_{cn} = \frac{2\pi EI_{11}}{n_sR^2} \left( \tilde{\mu}\bar{n}^4 + \frac{\lambda}{\bar{n}^2}\right)
	\end{equation}
Minimizing Eqn. \ref{eq:Tcn_small_mu} with respect to $\bar{n}$ by setting the derivative $dT_{cn}/d\bar{n}=0$, we obtain the scaling law
	\begin{equation}\label{eq:nbar_small_mu}
	\bar{n} = \left(\frac{\lambda}{2\tilde{\mu}} \right)^{1/6}
	\end{equation}
Substituting Eqn. \ref{eq:nbar_small_mu} into Eqn. \ref{eq:Tcn_small_mu} yields a power law for $T_c$.
	\begin{equation}\label{eq:Tc_small_mu}
	T_c \approx \frac{11.875}{n_sR^2} \, \left(\bar{k}_{uu}^{el}R^4 \right)^{2/3} (\widetilde{GJ})^{1/3}
	\end{equation}
Equation \ref{eq:Tc_small_mu} suggests that when the torsional stiffness is smaller than the bending stiffness, the buckling behavior is dominated by torsion and does not depend explicitly on the bending stiffness at all. Equation \ref{eq:Tc_small_mu} is shown in Fig. \ref{fig:Tc_nondim} next to Eqn. \ref{eq:Tc_nondim} and finite-element calculations. The power law is strictly less than Eqn. \ref{eq:Tc_nondim} which conveniently gives a conservative design criterion.

\subsection{Lateral stiffness of spoke system much greater than rim bending stiffness}
A second power law regime appears when $\lambda \gg 1$ and $\tilde{\mu} \sim 1$. Under these conditions, Eqn. \ref{eq:Tc_nondim} becomes
	\begin{equation}\label{eq:Tcn_boef}
	T_{cn} = \frac{2\pi EI_{11}}{n_sR^2}\left(\bar{n}^2 + \frac{\lambda}{\bar{n}^2} \right)
	\end{equation}
Following the same procedure in section \ref{sec:powerlaw_1} gives another separate power law:
	\begin{equation}\label{eq:Tc_boef}
	T_c \approx \frac{4\pi}{n_s} \left(\bar{k}_{uu}^{el}EI_{11} \right)^{1/2} 
	\end{equation}
Substituting Eqn. \ref{eq:TN} for $T_c$ gives $N_{rc}=2\left(\bar{k}_{uu}^{el}EI_{11}\right)^{1/2}$, which is precisely the buckling load for an infinite, straight beam on an elastic foundation~\cite{Hetenyi1946b}. Figure \ref{fig:Tc_nondim} (c) shows a comparison of Eqn. \ref{eq:Tc_boef} against Eqn. \ref{eq:Tc_nondim} and finite-element calculations.

@@

The applied lateral force and torque are determined using Castigliano's first theorem. For an applied force $P$ and no applied torque, we have
	\begin{equation}\label{eq:Castig}
	\begin{split}
	P &= \frac{\partial U_n}{\partial u_n}\\% = \frac{\partial^2 U_n}{\partial u_n^2}u_n + \frac{\partial^2 U_n}{\partial u_n \partial \phi_n}\phi_n\\
	0 &= \frac{\partial U_n}{\partial \phi_n}% = \frac{\partial^2 U_n}{\partial \phi_n^2}\phi_n + \frac{\partial^2 U_n}{\partial \phi_n \partial u_n}u_n
	\end{split}
	\end{equation}
Substituting $U_{rim} + U_{spokes}$ for $U_n$ and inserting the mode shape Eqn. \ref{eq:modeshape}, Eqn. \ref{eq:Castig} becomes a linear system of equations for $u_n$ and $\phi_n$. Solving \ref{eq:Castig} for $u_n$ yields
	\begin{equation}\label{eq:un}
	u_n = \left(\frac{PR^3}{\pi}\right) \frac{EI_{11}+\widetilde{GJ}n^2}{EI_{11}\widetilde{GJ}n^2(n^2-1)^2 + R\bar{k}_{uu}(EI_{11}+\widetilde{GJ}n^2)}
	\end{equation}
Defining an equivalent torsion stiffness, bending stiffness, and spoke stiffness, the mode stiffness $P/u_n$ can be obtained from Eqn. \ref{eq:un} as
	\begin{equation}\label{eq:Kn_spring}
	K_n = \frac{K_{bending}K_{torsion}}{K_{bending}+K_{torsion}} + K_{spokes}
	\end{equation}
\,\,\,\, where
	\begin{equation}\label{eq:Ktorsion}
	K_{torsion} = \frac{\pi\widetilde{GJ}n^2}{R^3}(n^2-1)^2
	\end{equation}
	\begin{equation}\label{eq:Kbending}
	K_{bending} = \frac{\pi EI_{11}}{R^3}(n^2-1)^2
	\end{equation}
	\begin{equation}\label{eq:Kspokes}
	K_{spokes} = \pi R \bar{k}_{uu}^{el}
	\end{equation}
The form of Equation \ref{eq:Kn_spring} suggests an equivalent spring model (Fig. \ref{fig:Kn}) for the $n$th mode in which the rim stiffness is represented by two springs connected which is then connected in parallel with the spoke stiffness. If there is a large difference between the rim bending stiffness and torsion stiffness, the equivalent rim stiffness will be dominated by the smaller of the two.

@@

\bibliographystyle{naturemag}
\bibliography{C:/Users/matt/OneDrive/Documents/Research/bibtex/Papers-BMD2016.bib}

>>>>

\section*{ABSTRACT}
The spokes of a bicycle wheel brace and stiffen its slender rim, but only as long as the pre-tension is high enough to prevent them from going slack under external loads. However, if the spoke tension is too high, the rim will buckle into a non-planar shape. Despite its importance for wheel strength and safety, we know of no formula in the literature for the maximum tension a wheel can withstand before buckling.

Here we derive such a formula. The non-dimensionalized buckling tension depends only on two parameters which are identified as the ratio of rim torsional stiffness to rim bending stiffness, and the ratio of lateral spoke bracing stiffness to rim bending stiffness. The buckled rim adopts a sinusoidal lateral profile where the number of positive peaks $n\geq 2$ is determined by these two stiffness ratios. The lateral stiffness of the wheel may be represented by a system of equivalent springs: two series-connected springs for the rim bending and torsion stiffness, connected in parallel with a spring for the spoke system.

@@

He did not discuss the problem of buckling under tension further, but proposed that wheel failure commonly occurs due to lateral buckling when external loads have caused spokes to lose tension. Pippard and Francis~\cite{Pippard1932d} derived an equation for lateral deflection based on an elastic foundation model, but did not discuss stability and neglected the effect of spoke pre-tension.

The theory of bicycle wheel buckling is closely related to the theory of flexural-torsional buckling of rings and arches. Timoshenko and Gere~\cite{Timoshenko1961a} gave a formula for the critical load for a ring with doubly-symmetric cross-section subjected to radial loads. The theory of flexural-torsional buckling of monosymmetric arches (bicycle rim cross-sections generally have only one plane of symmetry) was broadly formalized by Trahair and Papangelis~\cite{Trahair1987b} using the virtual work approach to derive the equilibrium and stability equations. Their theory has been extended to treat arches restrained by continuous~\cite{Pi2002b} or discrete~\cite{Bradford2002d} elastic supports and elastic end restraints~\cite{Guo2014b}.

@@

The buckling tension is controlled by the geometric and material properties of the wheel. The wheel designer may wish to achieve the stiffest and strongest wheel for a given size and weight by tuning parameters such as spoke count, spoke angle, rim width and depth, and material. Here we will answer the question of precisely how each of these factors affects the maximum safe tension in order to guide rational wheel design and component selection.

@@

To hold the spokes in tension, the rim must be under compression. An accurate approximation\footnote{95\% accurate for 6 spokes, $\alpha=10^{\circ}$, $\beta=10^{\circ}$, where angles $\alpha$ and $\beta$ are defined in Fig. \ref{fig:schematic}. Increasing the number of spokes only improves the estimate.} of the compressive force was given by Sharp~\cite{Sharp1977a} by considering a force balance on the top (or bottom) half of the rim.
	\begin{equation}\label{eq:TN}
	N_r = \frac{n_sT}{2\pi}
	\end{equation}
The spokes act like guy-wires, both supporting the rim radially and stabilizing the rim laterally. However, if the spoke tension exceeds a critical value $T_c$, the compression in the rim causes the wheel to buckle laterally, bending and twisting out of its initial plane, as shown in Fig. \ref{fig:schematic} (c).

@@

The deformed configuration of the rim is described by the displacement of the shear center $\bm{u}=u\e + v\ee + w\eee$ and the twist angle $\phi$. The in-plane curvature $\kappa_1$ remains constant during flexural-torsional buckling. The in-plane curvature, out-of-plane curvature and twist rate are
	\begin{equation}\label{eq:k1}
	\kappa_1 = 1/R
	\end{equation}
	\begin{equation}\label{eq:k2}
	\kappa_2 = u'' - \frac{\phi}{R}
	\end{equation}
	\begin{equation}\label{eq:k3}
	\kappa_3 = \phi' + \frac{1}{R} u'
	\end{equation}
The symbol $()'$ indicates a derivative with respect to arc-length $s$ along the \eee{} direction. It can easily be verified that Equations \ref{eq:k1}--\ref{eq:k3} simplify to the standard equations for straight beams in the limit $R\rightarrow \infty$. The total strain energy is decomposed into lateral bending, uniform torsion, and warping terms which are related to the curvatures.
% The curvature and twist rate produce a bending moment and twisting moment equal to
	% \begin{equation}\label{eq:M2}
	% M_2 = EI_{11} \kappa_2
	% \end{equation}
	% \begin{equation}\label{eq:M3}
	% M_3 = GJ \kappa_3 + EI_w \kappa_3''
	% \end{equation}
	\begin{equation}\label{eq:Urim}
	U_{rim} = \frac{1}{2} \int_0^{2\pi} \left( EI_{11} \kappa_2^2 + GJ \kappa_3^2 + EI_w (\kappa_3')^2 \right)\, ds
	\end{equation}
The warping energy $EI_w(\kappa_3')^2$ is primarily related to the bending energy in the rim sidewalls (which act as I-beam flanges) of the rim due to twisting\footnote{See http://www.steel-insdag.org/TeachingMaterial/Chapter17.pdf for a simple treatment.}. For a single-wall rim, the energy contribution from warping can exceed the contribution from torsion.

@@

In this paper we consider bicycle wheels with slender prestressed spokes. We model a single spoke as an elastic bar pinned at each end, ensuring that the spoke force is always collinear with the spoke axis. The hub is assumed to be rigid. The linearized elongation of a spoke is given by
	\begin{equation}\label{eq:selong}
	\Delta l = \bm{u}_n\cdot \n
	\end{equation}
where $\bm{u}_n$ is the displacement vector of the spoke nipple and \n{} is a unit vector pointing along the axis of the spoke. The linearized rotation of the spoke is given by
	\begin{equation}\label{eq:srot1}
	\Omega_1 = \frac{1}{l} \bm{u}_n\cdot \nn
	\end{equation}
	\begin{equation}\label{eq:srot2}
	\Omega_2 = \frac{1}{l} \bm{u}_n\cdot \nnn
	\end{equation}
where \nn{} and \nnn{} are two orthogonal unit vectors orthogonal to \n{}, and $l$ is the spoke length. The elongation produces a net force on the rim parallel to the original spoke axis equal to
	\begin{equation}\label{eq:sF1}
	f_1 = E_sA_s\left(\frac{\Delta l}{l}\right)\n = \frac{E_sA_s}{l} \n \cdot \bm{u}_n 
	\end{equation}
The rotation of the spoke produces net forces on the rim perpendicular to the original spoke axis due to the rotation of the initial tension vector.
	\begin{equation}\label{eq:sF2}
	f_2 = T \sin{\Omega_1} \approx \frac{T}{l} \nn \cdot \bm{u}_n
	\end{equation}
	\begin{equation}\label{eq:sF3}
	f_3 = T \sin{\Omega_2} \approx \frac{T}{l} \nnn \cdot \bm{u}_n
	\end{equation}
Since \n{}, \nn{}, \nnn{} are mutually orthogonal, the total force on the rim is given by $\bm{f} = \bm{k}_f \cdot \bm{u}_n$, where the spoke stiffness matrix is given by
	\begin{equation}\label{eq:kf}
	\bm{k}_f = \frac{E_sA_s}{l}\n\n + \frac{T}{l}(\nn\nn + \nnn\nnn)
	\end{equation}
The first term in Eqn.~\ref{eq:kf} is the {\bf elastic stiffness} and the second term is the {\bf tension stiffness}. The tension stiffness is the component responsible for the transverse vibrations of a guitar string, e.g. The term $\n\n$ is the dyadic product, or tensor product (also called the outer product), and is defined by the property $\n\n\cdot \bm{a} = (\n\cdot \bm{a})\n$ for any vector $\bm{a}$.

@@

\subsection{Total potential energy and buckling criterion}
The wheel becomes unstable when the increase in strain energy is exactly balanced by the virtual work for a perturbing displacement $\delta u, \delta\phi$, i.e. when the total potential energy is zero.
	\begin{equation}\label{eq:TotPot}
	\Pi = U_{rim} + U_{spokes} - V_{rim} = 0
	\end{equation}
We assume a buckled shape of the form
	\begin{equation}\label{eq:modeshape}
	\begin{split}
	u &= \delta u_n \cos{\left(\frac{ns}{R}\right)}\\
	\phi &= \delta\phi_n \cos{\left(\frac{ns}{R} \right)}
	\end{split}
	\end{equation}
\,\,\,\, where $n$ is the number of positive peaks of the lateral displacement. Substituting the buckling mode shape Eqn. \ref{eq:modeshape} into Eqns. \ref{eq:Urim}, \ref{eq:Vrim}, and \ref{eq:Us} and substituting into Eqn. \ref{eq:TotPot} yields a buckling criterion for the $n$th mode.
	\begin{multline}\label{eq:ECrit}
	\frac{\pi EI_{11}}{R^3}(n^2 \delta u_n - R\delta\phi_n)^2 + \frac{\pi n^2}{R^2}\left(GJ + \frac{EI_w}{R^2}n^2\right)(\delta u_n-R\delta\phi_n)^2 + \\
	\pi R(\bar{k}_{uu}^{el} + \bar{k}_{uu}^{el})\delta u_n^2 - N_r\frac{\pi n^2}{R}\delta u_n^2=0
	\end{multline}
The torsional stiffness and warping stiffness can be combined into an effective torsional stiffness $\widetilde{GJ} = GJ + n^2EI_w/R^2$. Equation \ref{eq:ECrit} has a quadratic form and can be written in matrix notation:
	\begin{equation}\label{eq:Qform}
	\begin{bmatrix}
	\delta u_n\\\delta\phi_n
	\end{bmatrix} \cdot
	\begin{bmatrix}
	\frac{\partial^2 \Pi}{\partial u_n^2} & \frac{\partial^2 \Pi}{\partial u_n \partial \phi_n}\\
	\frac{\partial^2 \Pi}{\partial u_n\partial\phi_n} & \frac{\partial^2 \Pi}{\partial \phi_n^2}
	\end{bmatrix} \cdot
	\begin{bmatrix}
	\delta u_n\\\delta\phi_n
	\end{bmatrix}
	=0
	\end{equation}
Non-trivial solutions of Eqn. \ref{eq:Qform} exist when the determinant of the matrix vanishes. Expanding the determinant in Eqn. \ref{eq:Qform} and substituting Eqns. \ref{eq:TN} and \ref{eq:kuu} yields a linear equation for the critical buckling tension of the $n$th mode.
	\begin{multline}\label{eq:Tc_eq}
	\left(\frac{2\pi EI_{11}}{R}+\frac{2\pi \widetilde{GJ}n^2}{R} \right) \left( \frac{2\pi \widetilde{GJ}n^2}{R^3} + \frac{2\pi EI_{11}n^4}{R^3} + 2\pi R\bar{k}_{uu}^{el} + \frac{n_sT_{cn}}{l} - \frac{n_sn^2T_{cn}}{R}\right)\\
	- \left( \frac{2\pi EI_{11}n^2}{R^2} + \frac{2\pi \widetilde{GJ}n^2}{R^2} \right)^2 = 0
	\end{multline}
Solving Eqn. \ref{eq:Tc_eq} for $T_{cn}$ yields the tension at which the $n$th mode becomes unstable. The critical buckling tension $T_c$ is the minimum $T_{cn}$ with respect to the mode number $n\geq 2$.

\section{BUCKLING DUE TO SPOKE TENSION}
The solution to Eqn. \ref{eq:Tc_eq} can be written in non-dimensional form as
	\begin{equation}\label{eq:Tc_nondim}
	\frac{T_{cn}}{T_e} = \left( \frac{\tilde{u}n^2(n^2-1)^2}{1+\tilde{\mu}n^2} + \lambda \right) \left(\frac{1}{n^2-R/l} \right)
	\end{equation}
\,\,\,\, where the ``Euler tension'' $T_e=2\pi EI_{11}/n_sR^2$ is the tension that would produce a compressive force in the rim equal to the classical buckling load for a straight beam-column of length $2\pi R$. The parameter $\tilde{\mu} = \widetilde{GJ}/EI_{11}$ is the ratio of the effective torsional stiffness to lateral bending stiffness. The parameter $\lambda = \bar{k}_{uu}^{el}R^4/EI_{11}$ is the ratio of the spoke system stiffness $\bar{k}_{uu}^{el}R$ to the rim bending stiffness $EI_{11}/R^3$. For most wheels, $R/l\approx 1$.

@@

\subsection{Rims with low torsional stiffness}\label{sec:powerlaw_1}
Many bicycle rims are built from thin-walled open channel sections which have very low torsional stiffness $\widetilde{GJ}$ compared with their bending stiffness. Replacing the discrete mode number $n$ with a continuous variable $\bar{n}$, we adopt the following ansatz: (a) $\bar{n}^2 \gg 1$ and (b) $\tilde{\mu}\bar{n}^2 \ll 1$. Under these conditions, Eqn. \ref{eq:Tc_nondim} becomes
	\begin{equation}\label{eq:Tcn_small_mu}
	T_{cn} = \frac{2\pi EI_{11}}{n_sR^2} \left( \tilde{\mu}\bar{n}^4 + \frac{\lambda}{\bar{n}^2}\right)
	\end{equation}
Minimizing Eqn. \ref{eq:Tcn_small_mu} with respect to $\bar{n}$ by setting the derivative $dT_{cn}/d\bar{n}=0$, we obtain the scaling law
	\begin{equation}\label{eq:nbar_small_mu}
	\bar{n} = \left(\frac{\lambda}{2\tilde{\mu}} \right)^{1/6}
	\end{equation}
Substituting Eqn. \ref{eq:nbar_small_mu} into Eqn. \ref{eq:Tcn_small_mu} yields a power law for $T_c$.
	\begin{equation}\label{eq:Tc_small_mu}
	T_c \approx \frac{11.875}{n_sR^2} \, \left(\bar{k}_{uu}^{el}R^4 \right)^{2/3} (\widetilde{GJ})^{1/3}
	\end{equation}
Equation \ref{eq:Tc_small_mu} suggests that when the torsional stiffness is smaller than the bending stiffness, the buckling behavior is dominated by torsion and does not depend explicitly on the bending stiffness at all. Equation \ref{eq:Tc_small_mu} is shown in Fig. \ref{fig:Tc_nondim} next to Eqn. \ref{eq:Tc_nondim} and finite-element calculations. For $\tilde{\mu}<0.13$, the power law is strictly less than Eqn. \ref{eq:Tc_nondim} which gives a conservative design criterion.

\subsection{Lateral stiffness of spoke system much greater than rim bending stiffness}
A second power law regime appears when $\lambda \gg 1$ and $\tilde{\mu} \sim 1$. Under these conditions, Eqn. \ref{eq:Tc_nondim} becomes
	\begin{equation}\label{eq:Tcn_boef}
	T_{cn} = \frac{2\pi EI_{11}}{n_sR^2}\left(\bar{n}^2 + \frac{\lambda}{\bar{n}^2} \right)
	\end{equation}
Following the same procedure in section \ref{sec:powerlaw_1} gives another power law:
	\begin{equation}\label{eq:Tc_boef}
	T_c \approx \frac{4\pi}{n_s} \left(\bar{k}_{uu}^{el}EI_{11} \right)^{1/2} 
	\end{equation}
Substituting Eqn. \ref{eq:TN} for $T_c$ gives $N_{rc}=2\left(\bar{k}_{uu}^{el}EI_{11}\right)^{1/2}$, which is precisely the buckling load for an infinite, straight beam on an elastic foundation~\cite{Hetenyi1946b}. Figure \ref{fig:Tc_nondim} (c) shows a comparison of Eqn. \ref{eq:Tc_boef} against Eqn. \ref{eq:Tc_nondim} and finite-element calculations.

@@

The applied lateral force and torque are determined using Castigliano's first theorem. For an applied force $P$ and no applied torque, we have
	\begin{equation}\label{eq:Castig}
	\begin{split}
	P &= \frac{\partial U_n}{\partial u_n}\\% = \frac{\partial^2 U_n}{\partial u_n^2}u_n + \frac{\partial^2 U_n}{\partial u_n \partial \phi_n}\phi_n\\
	0 &= \frac{\partial U_n}{\partial \phi_n}% = \frac{\partial^2 U_n}{\partial \phi_n^2}\phi_n + \frac{\partial^2 U_n}{\partial \phi_n \partial u_n}u_n
	\end{split}
	\end{equation}
Substituting $U_{rim} + U_{spokes}$ for $U_n$ and inserting the mode shape Eqn. \ref{eq:modeshape}, Eqn. \ref{eq:Castig} becomes a linear system of equations for $u_n$ and $\phi_n$. Solving \ref{eq:Castig} for $u_n$ yields
	\begin{equation}\label{eq:un}
	u_n = \left(\frac{PR^3}{\pi}\right) \frac{EI_{11}+\widetilde{GJ}n^2}{EI_{11}\widetilde{GJ}n^2(n^2-1)^2 + R\bar{k}_{uu}(EI_{11}+\widetilde{GJ}n^2)}
	\end{equation}
Defining an equivalent torsion stiffness, bending stiffness, and spoke stiffness, the mode stiffness $P/u_n$ can be obtained from Eqn. \ref{eq:un} as
	\begin{equation}\label{eq:Kn_spring}
	K_n = \frac{K_{bending}K_{torsion}}{K_{bending}+K_{torsion}} + K_{spokes}
	\end{equation}
\,\,\,\, where
	\begin{equation}\label{eq:Ktorsion}
	K_{torsion} = \frac{\pi\widetilde{GJ}n^2}{R^3}(n^2-1)^2
	\end{equation}
	\begin{equation}\label{eq:Kbending}
	K_{bending} = \frac{\pi EI_{11}}{R^3}(n^2-1)^2
	\end{equation}
	\begin{equation}\label{eq:Kspokes}
	K_{spokes} = \pi R \bar{k}_{uu}^{el}
	\end{equation}
Equation \ref{eq:Kn_spring} corresponds to the equivalent spring model (Fig. \ref{fig:Kn}) for the $n$th mode. Since the rim bending stiffness are connected in series, the equivalent rim stiffness will be dominated by the smaller of the two.

@@

\normalem
\bibliographystyle{naturemag}
\bibliography{C:/Users/matt/OneDrive/Documents/Research/bibtex/Papers-BMD2016.bib}
